Generalized Tractability

### 2023-04-21 GeneralizedTractability_230418.tex
Fred's Home Computer

\fred{We need to check the proof for
\begin{itemize}
\item correctness
\item exponents
\item whether the additional assumptions are too restrictive and can be generalized
\item whether we need $\tau$ in the assumption
\item Do we need non-decreasing or strictly increasing?
\end{itemize}}
\begin{definition}[Weak Tractability]
If $n(\varepsilon,d)$ denotes the minimal number of function values needed to approximate all functions. We say that the problem is \emph{weakly tractable} if
    \[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\ln n(\varepsilon,d)}{\varepsilon^{-1}+d} = 0\]
\end{definition}
%\begin{definition}[Quasi-polynomial Tractability]
 %   The problem is \emph{quasi-polynomially tractable}, if there exist two constants $C, t> 0$ such that
  %  \[
   % n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}
    %\] for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.
%\end{definition}

\peter{
\section{Weak $T$-$(s,t)$-weak tractability} \label{sec:weak}

\begin{definition}[$T$-$(s,t)$-weak tractability]
Let $s,t>0$. We say that the problem is \emph{$T$-$(s,t)$ weakly tractable} if
    \[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\log n(\varepsilon,d)}{T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp)} = 0.\]
\end{definition}

We show a necessary and sufficient condition for $T$-$(s,t)$-weak tractability, working analogously to \cite{WW17}.
\begin{theorem}\label{thm:weak}
For $c>0$, define
\[
\sigma (c,d,s,\vp):= \sum_{i=1}^\infty \exp \left(-c\, T(\lambda_{i,d}^{-s},1,\vp)\right).
\]
Then the problem is $T$-$(s,t)$-weakly tractable if and only if
\[
\mu (c,s,t,\vp)=\sup_{d\in\naturals} \sigma (c,d,s,\vp) \exp\left(-c\, T(0,d^t,\vp)\right)<\infty
\]
holds for all $c>0$.
\end{theorem}
\begin{proof}
 We first show the necessity of the condition in the theorem.   Indeed, suppose that the problem is  $T$-$(s,t)$-weakly tractable. Then, for any $c>0$ there exists a positive integer $V=V(c)$ such that  
 \[
 n(\varepsilon,d)\le 
 \left\lfloor \exp\left(c\,
 \left(T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp)\right)
 \right) \right\rfloor
 \]
 whenever $\varepsilon^{-1} + d \ge V$. For $d\in\naturals$ 
 choose $\varepsilon>0$ such that $\varepsilon^{-1}\ge \max\{1,V-d\}$. Since the eigenvalues $\lambda{i,d}$ are non-increasing in $i$, we know that
 \[
 \lambda_{\lfloor \exp(c\,
 (T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp))) \rfloor+1}^2 \le \varepsilon^2
 \]
 for any $\varepsilon$ satisfying $\varepsilon^{-1}\ge \max\{1,V-d\}$.
Let 
\[
n=n(\varepsilon):= \lfloor \exp(c\,
 (T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp))) \rfloor+1
\]
and
\[
n^*:= \lfloor \exp(c\,
 (T(\max\{1,(V-d)^s\},1,\vp) + T(0,d^t,\vp))) \rfloor+1.
\]
By varying $\varepsilon^{-1}$ in the interval 
$[\max\{1,V-d\},\infty)$ we can make $n(\varepsilon)$ attain 
any integer value greater than or equal to $n^*$.

Furthermore, we have, as $\lambda_{n,d}\le \varepsilon$ and as $T$ is monotonically increasing, 
\begin{eqnarray*}
n(\varepsilon)-1 & \le & \exp(c\, (T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp)))\\
& \le & \exp(c\, (T(\lambda_{n(\varepsilon),d}^{-s},1,\vp) + T(0,d^t,\vp))),
\end{eqnarray*}
which is equivalent to 
\[
\exp(-2c\, (T(\lambda_{n(\varepsilon),d}^{-s},1,\vp) + T(0,d^t,\vp))) \le \frac{1}{(n(\varepsilon)-1)^2}.
\]
Therefore,
\begin{eqnarray*}
    \exp(-2c\, T(0,d^t,\vp)) \sum_{i=1}^\infty \exp(-2c\, T(\lambda_{i,d}^{-s},1,\vp))
    &\le& \exp(-2c\, T(0,d^t,\vp))\, (n^*-1) + \sum_{i=n^*}^\infty \frac{1}{(i-1)^2}\\
    &\le& \exp(-2c\, T(0,d^t,\vp))\, (n^*-1) + \frac{\pi^2}{6}.
\end{eqnarray*}
Moreover,
\begin{eqnarray*}
    \exp(-2c\, T(0,d^t,\vp))\, (n^*-1) &\le & \exp(-2c\, T(0,d^t,\vp))\, 
    \exp (c\, T(V^s,1,\vp) + c\, T(0,d^t,\vp))\\
    &\le & \exp (c\, T(V^s,1,\vp))\\
    &<& \infty,
\end{eqnarray*}
independently of $d$. This implies that 
\[
\exp(-2c\, T(0,d^t,\vp)) \sum_{i=1}^\infty \exp(-2c\, T(\lambda_{i,d}^{-s},1,\vp))<\infty,
\]
also independently of $d$. As $c>0$ was chosen arbitrarily, we can replace $2c$ by $c$ and see
necessity of the condition in the theorem.

We now show sufficiency.

Suppose that $\mu(c,s,t,\vp) <\infty$ for all $c>0$. It is no loss of generality to assume that 
$\mu(c,s,t,\vp) \ge 1$. Since the terms of the series in the definition of $\mu(c,s,t,\vp) $ are non-increasing, we have 
\[
\exp (-c\, T(0,d^t,\vp))\, n\, \exp (-c\, T(\lambda_{n,d}^{-s},1,\vp)) \le \mu (c,s,t,\vp).
\]
Then we obtain 
\[
\exp (-c\, T(\lambda_{n,d}^{-s},1,\vp)) \le \frac{1}{n}\, \mu (c,s,t,\vp)\, 
\exp (c\, T(0,d^t,\vp)),
\]
which is equivalent to 
\[
\exp (c\, T(\lambda_{n,d}^{-s},1,\vp)) \ge \frac{n}{\mu (c,s,t,\vp)\, 
\exp (c\, T(0,d^t,\vp))},
\]
which in turn is equivalent to 
\[
T(\lambda_{n,d}^{-s},1,\vp) \ge \frac{1}{c}\log \left(\frac{n}{\mu (c,s,t,\vp)\, 
\exp (c\, T(0,d^t,\vp))}\right)
\]
as long as 
\[
n> \mu (c,s,t,\vp)\, \exp (c\, T(0,d^t,\vp)).
\]
This implies that for the special choice 
\[
n= \left\lceil 
   \mu (c,s,t,\vp)\, \exp \left(c\, (T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp))\right)
   \right\rceil 
\]
we have
\[
T(\lambda_{n,d}^{-s},1,\vp) \ge T(\varepsilon^{-1},1,\vp),
\]
which implies that $\lambda_{n,d} \le \varepsilon$, and thus $n(\varepsilon,d)\le n$. Note that the 
argument $x$ of the ceiling function in the definition of our special $n$ is at least one, which implies 
$\lceil x \rceil \le 2x$.

Consequently,
\[
n(\varepsilon, d) \le \exp\left(c\, \left(
\frac{\log (2\mu (c,s,t,\vp) )}{c}
T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp)
\right)\right).
\]
Clearly, there must exist a positive $V=V(c,s,t,\vp)$ such that when $\varepsilon^{-1} +d \ge V$ it holds true that
\[
  T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp) \ge \frac{\log (2\mu (c,s,t,\vp) )}{c}.
\]
Hence we finally have 
\[
n(\varepsilon,d) \le \exp (2c\, (T(\varepsilon^{-s},1,\vp) + T(0,d^t,\vp))).
\]
Since the latter inequality holds for every $c>0$, we have $T$-$(s,t)$-weak tractability.

\end{proof}
}

@@

\noindent \textbf{Sufficient Condition:}\\	
Suppose that $S_c <\infty$ for all $c>0$.  Since the terms of the series in the definition of $S_c$ are non-increasing, we have 
	\[
		\frac{n}{h(cT(\lambda_{n,d}^{-1},d))}\le S_c \qquad \forall n,d \in \naturals.
	\]
or equivalently
\[
	n\le S_c \, h(cT(\lambda_{n,d}^{-1},d))   \qquad \forall n,d \in \naturals.
\]
This implies that for the special choice 
\[
	n = \left\lceil 
	S_c \, h(cT(\varepsilon^{-1},d))  
	\right\rceil \ge 1
\]
	we have
	\[
	T(\lambda_{n,d}^{-s},d) \ge T(\varepsilon^{-1},d).
	\]
Since $T(\cdot,d)$ is non-decreasing, it follows that $\lambda_{n,d} \le \varepsilon$, and thus 
\[
\comp(\varepsilon,d) \le  \left\lceil 
S_c \, h(cT(\varepsilon^{-1},d))  
\right\rceil \qquad \forall c > 0.  
\]

\peter{
We could go on like this:

It is no loss of generality to assume that $S_c \ge 1$. From above we get 
\[
h^{-1}(\comp(\varepsilon,d)) \le h^{-1} \left( \left\lceil 
S_c \, h(cT(\varepsilon^{-1},d))  
\right\rceil \right) \le h^{-1} \left(  
2 S_c \, h(cT(\varepsilon^{-1},d))  
 \right) \qquad \forall c > 0,
\]
where we used that the argument of the ceiling function is at least one.

From this we deduce, by \eqref{eq:submult_h_inverse},
\[
h^{-1}(\comp(\varepsilon,d)) \le
 h^{-1} (2 S_c) + cT (\varepsilon^{-1},d).
\]
Now, for sufficiently large $\varepsilon^{-1}+d$, we must have
\[
c T (\varepsilon^{-1},d) \ge h^{-1} (2 S_c).
\]
This is because the finiteness of $S_c$ implies that $\lim_{\varepsilon \to 0} T(\varepsilon^{-1},d) = \infty$ for all $d \in \naturals$,  together with the hypothesis on $\lim_{d \to \infty} T(\varepsilon^{-1},d)$.
So, 
\[
h^{-1}(\comp(\varepsilon,d)) \le 2 cT (\varepsilon^{-1},d).
\]
Since the latter holds for any $c>0$, we have sub-$h\circ T$-tractability.

}

Consequently, choosing $c=1/2$, it follows that 
\begin{align}
\nonumber
\lim_{\varepsilon^{-1} + d \to \infty} \frac{\comp(\varepsilon,d)}{ h(T(\varepsilon^{-1},d))  }
& = \lim_{\varepsilon^{-1} + d \to \infty} \frac{\comp(\varepsilon,d)}{ h(2cT(\varepsilon^{-1},d))  } \\
& \le \lim_{\varepsilon^{-1} + d \to \infty} \frac{ \left\lceil 
	S_c \, h(cT(\varepsilon^{-1},d)) \right\rceil}{ [h(cT(\varepsilon^{-1},d))]^2  } \qquad \text{by \eqref{eq:h_cond}}
\\
\nonumber
& \le \lim_{\varepsilon^{-1} + d \to \infty} \frac{
	S_c \, h(cT(\varepsilon^{-1},d))+1}{ [h(cT(\varepsilon^{-1},d))]^2  }  \\
& = \lim_{\varepsilon^{-1} + d \to \infty} \left \{\frac{
	S_c }{ h(cT(\varepsilon^{-1},d))  }  + \frac{
	1}{ [h(cT(\varepsilon^{-1},d))]^2 } \right \} \label{eq:alastone}.
\end{align}	
The finiteness of $S_c$ implies that $\lim_{\varepsilon \to 0} T(\varepsilon^{-1},d) = \infty$ for all $d \in \naturals$.  Together with the hypothesis on $\lim_{d \to \infty} T(\varepsilon^{-1},d)$, we can infer that $\lim_{\varepsilon^{-1} + d \to \infty} T(\varepsilon^{-1},d) = \infty$.  This implies that the limit in \eqref{eq:alastone} vanishes.

@@

\noindent \textbf{Necessary Condition:}\\
Suppose that the problem is  sub-$h \circ T$ tractable. Then, for any $c>0$ there exists a positive integer $V_c$ such that  \fred{having a little trouble removing the $c$ here.} \peter{Here we do not need to remove the $c$, and actually shouldn't. Just in the limit relation it was not necessary.}
\[
	\comp(\varepsilon,d)\le 
	\left\lfloor h(cT(\varepsilon^{-1},d)) \right\rfloor \qquad \forall \varepsilon^{-1} + d \ge V_c.
\]
By the definition of the information complexity in \eqref{eq:comp_def}, it follows that 
\[
	\lambda_{n(\varepsilon,d,c),d} \le \varepsilon \quad \forall \varepsilon^{-1} + d \ge V_c, \qquad \text{where }
	n(\varepsilon,d,c):=\left\lfloor h(cT(\varepsilon^{-1},d)) \right\rfloor +1.
\]
Since $h(cT(\cdot,d))$ is non-decreasing, it follows that
\begin{equation}
		n(\varepsilon,d,c)\le  h(cT(\varepsilon^{-1},d)) +1\le  h(cT(\lambda^{-1}_{n(\varepsilon,d,c),d},d)) +1\qquad  \quad \forall \varepsilon^{-1} + d \ge V_c. \label{eq:nedcbound}
\end{equation}

@@

If $f(x)$ and $g(x)$ are both positive, and $\lim_{x \to \infty} g(x) = \infty$, is it true that
\[
\lim_{x \to \infty} \frac{f(x)}{g(x)} = 0 \iff 
\lim_{x \to \infty} \frac{\log(f(x))}{\log(g(x))} = 0?
\]

>>>>

%\begin{definition}[Quasi-polynomial Tractability]
 %   The problem is \emph{quasi-polynomially tractable}, if there exist two constants $C, t> 0$ such that
  %  \[
   % n(\varepsilon,d) \leq C\exp\{t(1+\ln(\varepsilon^{-1}))(1+\ln(d))\}
    %\] for all $\varepsilon \in (0,1)$ and $d\in\mathbb{N}$.
%\end{definition}

@@

\noindent \textbf{Sufficient Condition:}\\	
Suppose that $S_c <\infty$ for all $c>0$.  Since the terms of the series in the definition of $S_c$ are non-increasing, we have 
	\[
		\frac{n}{h(cT(\lambda_{n,d}^{-1},d))}\le S_c \qquad \forall n,d \in \naturals.
	\]
or equivalently
\[
	n\le S_c \, h(cT(\lambda_{n,d}^{-1},d))   \qquad \forall n,d \in \naturals.
\]
This implies that for the special choice 
\[
	n = \left\lceil 
	S_c \, h(cT(\varepsilon^{-1},d))  
	\right\rceil \ge 1
\]
	we have
	\[
	T(\lambda_{n,d}^{-s},d) \ge T(\varepsilon^{-1},d) \qquad \forall \varepsilon > 0, \, d \in \naturals.
	\]
Since $T(\cdot,d)$ is non-decreasing, it follows that $\lambda_{n,d} \le \varepsilon$, and thus 
\[
\comp(\varepsilon,d) \le  \left\lceil 
S_c \, h(cT(\varepsilon^{-1},d))  
\right\rceil \le   \lceil \underbrace{
S_c \, h(cT(0,1))}_{=: \widetilde{S}_c/2} \, h(cT(\varepsilon^{-1},d))  
\rceil\qquad \forall c, \varepsilon > 0, \, d \in \naturals.
\]

Since $S_c \ge 1/h(cT(0,1))$ by definition, it follows that $S_c \, h(cT(0,1)) \, h(cT(\varepsilon^{-1},d))  \ge 1$ and so 
\begin{align*}
\max(1,\comp(\varepsilon,d)) 
& \le  \lceil  S_c \, h(cT(0,1)) \, h(cT(\varepsilon^{-1},d))   \rceil 
 =  \lceil 
(\widetilde{S}_c/2) \, h(cT(\varepsilon^{-1},d))  
\rceil \\
& \le   \widetilde{S}_c \, h(cT(\varepsilon^{-1},d))  
\qquad\forall c, \varepsilon > 0, \, d \in \naturals.
\end{align*}
Since $h$ is strictly increasing, so is $h^{-1}$, and thus
\begin{align*}
h^{-1}(\max(1,\comp(\varepsilon,d)) ) & \le  h^{-1} \left(  
\widetilde{S}_c \, h(cT(\varepsilon^{-1},d))  \right) 
\\
&  \le  h^{-1} (2 S_c) + cT (\varepsilon^{-1},d).
\quad \forall c, \varepsilon > 0, \, d \in \naturals \qquad \text{by \eqref{eq:submult_h_inverse}.}
\end{align*}

The finiteness of $S_c$ implies that $\lim_{\varepsilon \to 0} T(\varepsilon^{-1},d) = \infty$ for all $d \in \naturals$.  Together with  the hypothesis on $\lim_{d \to \infty} T(\varepsilon^{-1},d)$ in \eqref{eq:h_cond}, it follows that $\lim_{\varepsilon^{-1} + d \to \infty} T(\varepsilon^{-1},d) = \infty$.  Therefore
\begin{equation*}
	\lim_{\varepsilon^{-1} + d \to \infty} \frac{h^{-1}(\max(1,\comp(\varepsilon,d)))}{T(\varepsilon^{-1},d)} 
	\le \lim_{\varepsilon^{-1} + d \to \infty} \frac{h^{-1} (2 S_c)}{T(\varepsilon^{-1},d)}  + c = c \qquad\forall c >  0.
\end{equation*}
Since this limit is bounded above by all positive $c$, it must be zero, and the problem is sub-$h$-$T$ tractable.

@@

\noindent \textbf{Necessary Condition:}\\
Suppose that the problem is  sub-$h \circ T$ tractable. Then, for any $c>0$ there exists a positive integer $V_c$ such that  
\begin{gather*}
	h^{-1}(\comp(\varepsilon,d))\le h^{-1}(\max(1,\comp(\varepsilon,d))) \le cT(\varepsilon^{-1},d)  \qquad \forall \varepsilon^{-1} + d \ge V_c, \\
	\comp(\varepsilon,d) \le \left\lfloor h(cT(\varepsilon^{-1},d)) \right\rfloor \qquad \forall \varepsilon^{-1} + d \ge V_c.
\end{gather*}
By the definition of the information complexity in \eqref{eq:comp_def}, it follows that 
\[
	\lambda_{n(\varepsilon,d,c),d} \le \varepsilon \quad \forall \varepsilon^{-1} + d \ge V_c, \qquad \text{where }
	n(\varepsilon,d,c):=\left\lfloor h(cT(\varepsilon^{-1},d)) \right\rfloor +1.
\]
Since $h(cT(\cdot,d))$ is non-decreasing, it follows that
\begin{equation}
		n(\varepsilon,d,c)\le  h(cT(\varepsilon^{-1},d)) +1\le  h(cT(\lambda^{-1}_{n(\varepsilon,d,c),d},d)) +1\qquad  \quad \forall \varepsilon^{-1} + d \ge V_c. \label{eq:nedcbound}
\end{equation}

@@

\begin{example}[Weak Tractability]
	Suppose that 
	\[
	T(\varepsilon^{-1},d,\vp) = \varepsilon^{-s} + d^t, \quad \forall \varepsilon > 0, \ d \in \naturals, \quad \text{where } \vp = (s,t), \qquad h(x) = \exp(x) \quad \forall x \ge 0.
	\]
	Then Theorem \ref{thm:subhT} implies that 
	\[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\log( \max(1,\comp(\varepsilon,d)))}{\varepsilon^{-s}+d^t} = 0 
	\iff 
	 \sup_{d\in\naturals}  \, \sum_{i=1}^\infty \exp(-c\, (\lambda_{i,d}^{-s} + d^t )) <\infty \qquad \forall c > 0.
	\]
	The expression on the left is the notion of $(s,t)$-weak tractability \cite{??}, and the condition on the right was derived in \cite{??}.  Weak tractability corresponds to $(s,t)=(1,1)$, and uniform weak tractability corresponds to the case when a problem is weakly tractable for all positive $s$ and $t$.
\end{example}
